Supplemental Materials for ``Universals of word order reflect optimization of grammars for efficient communication''

### 2019-10-14 writeup/pnas-word-order-si.tex
code

Trees are linearized using one of two grammars: One (`Correlating') places \emph{obj}, \emph{xcomp}, and \emph{acl} dependents on the same side of the head, and (in accordance with crosslinguistic tendencies, which are predicted by DLM) places the \emph{obj} dependents closer to the head than \emph{xcomp} dependents.
The other grammar (`Anti-Correlating') places \emph{xcomp} and \emph{acl} dependents opposite to \emph{obj} dependents.

We formulated this model as a binary-branching PCFG, and used a CKY parser to estimate $I[\mathcal{T}, \mathcal{U}]$ from 10,000 random sample sentences.

@@

\begin{figure}[ht]
    \centering
{
\centering{
\begin{dependency}[theme = simple]
   \begin{deptext}[column sep=1em]
	   V \& N \& V \& N  \\
   \end{deptext}
   \depedge{1}{2}{obj}
   \depedge{1}{3}{xcomp}
   \depedge{3}{4}{obj}
\end{dependency}

\begin{dependency}[theme = simple]
   \begin{deptext}[column sep=1em]
	   N \& V \& N \& V  \\
   \end{deptext}
   \depedge{2}{1}{obj}
   \depedge{2}{4}{xcomp}
   \depedge{4}{3}{obj}
\end{dependency}

}
}
	\caption{An English sentence with annotated syntactic relations.}
	\label{fig:sent-dep}
\end{figure}

@@

>>>>

Trees are linearized using one of two grammars: One (`Correlating') places \emph{obj}, \emph{xcomp}, and \emph{acl} dependents on the same side of the head, and (in accordance with crosslinguistic tendencies) places the \emph{obj} dependents closer to the head than \emph{xcomp} dependents.
The other grammar (`Anti-Correlating') places \emph{xcomp} and \emph{acl} dependents opposite to \emph{obj} dependents.

An example is provided in Figure~\ref{fig:sent-dep}.
We show how the two grammars linearize the same syntactic dependency structure:
The correlating grammar (left) linearizes the three relation types towards the right of the head; the anti-correlating one places \emph{obj} dependencies on the left and the other dependencies on the right.
This example provides some intuitive idea of why the correlating grammar might lead to improved parseability:
Note that the red token labeled `N' occupies the same structural position in both versions.
In the anti-correlating version (right), when given only the token sequence, without the syntactic structure, this word could a priori be an \emph{obj} dependent of any of the three verbs occurring to its right.
In the correlating version (left), this `N' token can only possibly be a dependent of the verb occurring to its left.

\begin{figure}[ht]
    \centering
{
\centering{
\begin{dependency}[theme = simple]
   \begin{deptext}[column sep=1em]
	   V \& \textcolor{red}{N} \& V \& N \& V  \\
   \end{deptext}
   \depedge{1}{2}{obj}
   \depedge{2}{3}{acl}
   \depedge{3}{4}{obj}
   \depedge{1}{5}{xcomp}
\end{dependency}
\begin{dependency}[theme = simple]
   \begin{deptext}[column sep=1em]
	   \textcolor{red}{N} \& N \& V \&  V \&  V   \\
   \end{deptext}
   \depedge{4}{1}{obj}
   \depedge{1}{3}{acl}
   \depedge{3}{2}{obj}
   \depedge{4}{5}{xcomp}
\end{dependency}

}
}
	\caption{Linearizations of a syntactic dependency structure, under a correlating grammar (left) and an anti-correlating grammar (right).}
	\label{fig:sent-dep}
\end{figure}

In order to test this intuition on the level of the entire tree distribution, we formulated this model as a binary-branching PCFG, and used a CKY parser to estimate $I[\mathcal{T}, \mathcal{U}]$ from 10,000 random sample sentences.

@@

@@

This simulation shows that correlations can in principle improve parseability in the controlled setting of such a model, without appeal to memory limitations; we leave a full graph-theoretical understanding to future work.
